### 2012-10-01 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updating section headers
linear problem updates

\chapter{Conventional Solution Methods for Linear Problems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems.  In addition, these matrices
are typically sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
p  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

\section{Stationary Methods}
\label{sec:stationary_methods}
Stationary methods arise from splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable\footnote{We may generalize this to
  non-diagonalizable matrices with the Jordan canonical form of
  $\ve{H}$.} \citep{saad_iterative_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

@@

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. Furthermore, variants are available that reduce the number
of global communications required (consider
\cite{sosonkina_scalable_1998} as an example of reducing global
operation counts using a different orthogonalization procedure than
Arnoldi), however, we will only consider the basic algorithms here as
this handful operations can be generalized to fit more complicated
algorithms. In all of these cases, we assume a general matrix/vector
formulation that is distributed in parallel such that both local and
global knowledge of their decomposition is available on
request. Furthermore, it is assumed that these objects are partitioned
in such a way that the parallel formulation of the operator and
vectors in Eq~(\ref{eq:linear_problem}) will be such that each
parallel process contains only a subset of the global problem and that
subset forms a local set of complete equations. The form of this
partitioning is problem dependent and often has a geometric or
graph-based aspect to its construction in order to optimize
communication patterns. Libraries such as Zoltan
\citep{devine_zoltan_2002}, provide an implementations of such
algorithms.

@@

\subsection{Parallel Vector Product}
\label{subsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses two vectors to a single
value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.

>>>>

\chapter{Conventional Solution Methods for Linear Problems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems. In many cases, these
matrices are sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

\section{Stationary Methods}
\label{sec:stationary_methods}
Stationary methods for linear systems arise from splitting the
operator in Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable\footnote{We may generalize this to
  non-diagonalizable matrices with the Jordan canonical form of
  $\ve{H}$.} \citep{saad_iterative_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

@@

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. We note here that variants are available that reduce the
number of global communications required (consider
\cite{sosonkina_scalable_1998} as an example of reducing global
operation counts using a different orthogonalization procedure than
Arnoldi), however, we will only consider the basic algorithms here as
this handful of operations can be generalized to fit more complicated
algorithms. In all of these cases, we assume a general matrix/vector
formulation that is distributed in parallel such that both local and
global knowledge of their decomposition is available on
request. Furthermore, it is assumed that these objects are partitioned
in such a way that the parallel formulation of the operator and
vectors in Eq~(\ref{eq:linear_problem}) will be such that each
parallel process contains only a subset of the global problem and that
subset forms a local set of complete equations. The form of this
partitioning is problem dependent and often has a geometric or
graph-based aspect to its construction in order to optimize
communication patterns. Libraries such as Zoltan
\citep{devine_zoltan_2002}, provide an implementations of such
algorithms.

@@

\subsection{Parallel Vector Product}
\label{subsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses a set of vectors to a
single value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.
